\usemodule[abr-03]
\usemodule[tugboat]

\setupbodyfontenvironment[default][em=italic]
% I prefer emphasized text to be italic rather than slanted

\setvariables
  [tugboat]
  [type=article,
   columns=yes,
   grid=no]

%\setvariables
%   [tugboat]
%   [year=\tubyear,
%    volume=\tubvol,
%    number=\tubnum,
%    page=\tubpage]

\setvariables
  [tugboat]
  [   title={\CONTEXT\ basics for users:\\ Conditional processing},
     author={Aditya Mahajan},
    address={},
      email={adityam (at) umich dot edu},
  ]


\definealternativestyle [emph]     [\em]     []

\StartAbstract
Very often, you want to generate multiple versions of the same document:
one version for printing and one for viewing on the screen, one version for
students and one version for the instructor, and so on. You can do this in a
simple but naive way: create different files set up for the different versions and
\type{\input} the common material, or create some new conditional flags using
\type{\newif} and set them appropriately for conditional processing. Or you
could use \emph{modes}\Dash the \CONTEXT\ way of doing conditional processing.
\StopAbstract

\starttext
\StartArticle 

\section {Introduction}

A mode is similar to a conditional flag, but with a few advantages: new modes
need not be explicitly defined (no need for something like \type|\newif|),
multiple modes can be simultaneously enabled or disabled, and the status of multiple
modes can be checked easily. Moreover, modes can be set from a command line
switch. As a result, multiple versions of a document can be generated without changing
the source file.

The name or identifier of a mode can be any combination of letters, digits, or
spaces. Names starting with \type{*} are reserved for system modes. 

In this article I explain how to activate a mode and how to check if a mode is
active or not. 

\section {Setting modes}

\CONTEXT\ has three commands for setting modes:
\startitemize
  \item \type{\enablemode [...]}
  \item \type{\disablemode[...]}
  \item \type{\preventmode[...]}
\stopitemize
The names are self-descriptive. \type{\enablemode} activates a mode,
\type{\disablemode} deactivates a mode, and \type{\preventmode} permanently
deactivates a mode. All three commands take a list of modes as an argument. For
example,  you can  activate modes named \type{screen} and \type{solution} with
\starttyping
    \enablemode[screen,solution]
\stoptyping

Modes can also be activated by a command line switch \type|-|\type|-modes| to
\type{texexec} or \type{context}. For example, another way to activate the \type{screen} and
\type{solution} modes, to run \CONTEXT\ using one of:
\starttyping 
    texexec --mode=screen,solution ...
    context --mode=screen,solution ...
\stoptyping


\section {Conditional processing based on modes}

Suppose you want to change the paper size of a document depending on whether it
is for print or screen. This can be done in multiple ways. You could set
the default paper size for print and change it in screen mode:
\starttyping
  \setuppapersize[letter][letter]
  \startmode[screen]
    \setuppapersize[S6][S6]
  \stopmode
\stoptyping
\noindentation 
(S6 is one of the screen-optimized paper sizes in \CONTEXT; the paper size has a
4:3 aspect ratio and a width equal to the width of A4 paper.)

Alternatively, you
could set a default paper size for the screen and change it if screen mode is
not enabled:
\starttyping
  \setuppapersize[S6][S6]
  \startnotmode[screen]
    \setuppapersize[letter][letter]
  \stopnotmode
\stoptyping

\type|\startmode| and \type|\startnotmode| can check for multiple modes,
by giving a list of modes as their arguments.  \type|\startmode|
processes its contents (everything until the next
\type|\stopmode|\footnote{This means that \type|\startmode| cannot be
nested.})  if any of the modes are enabled, otherwise (i.e., when all
the modes are disabled) \type|\startmode| ignores its
contents. The opposite is \type|\startnotmode|: it processes its
contents (everything until the next \type|\stopnotmode|) if any of the
modes are disabled, otherwise\Dash when all the modes are enabled\Dash
the contents are ignored.

\type|\startmode| and \type|\startnotmode| are ``\emph{or}'' environments. They
process their contents if any of the modes satisfy the required condition. Their
``\emph{and}'' counterparts are also available: \type|\startallmodes| and \type|\startnotallmodes|
process their contents only if all the given modes satisfy the required
condition. For example, suppose you want to enable interaction (e.g., hyperlinks)
only when both \type|screen| and \type|solution| modes are enabled. Then you can
use:
\starttyping
  \startallmodes[screen,solution]
    \setupinteraction[state=start]
  \stopallmodes
\stoptyping

To summarize, the four start-stop environments for checking modes are:
\starttyping
\startmode[mode1, mode2, ...]
  % Processed if any of the modes is enabled
\stopmode

\startnotmode[mode1, mode2, ...]
  % Processed if any of the modes is disabled
\stopnotmode

\startallmodes[mode1, mode2, ...]
  % Processed if all the modes are enabled
\stopallmodes

\startnotallmodes[mode1, mode2, ...]
  % Processed if all the modes are disabled
\stopnotallmodes
\stoptyping

These environments have \type|\doif...| alternatives that are useful for short
setups.  Also, they can be nested.

\starttyping
\doifmode        {modes} {content}
\doifnotmode     {modes} {content}
\doifallmodes    {modes} {content}
\doifnotallmodes {modes} {content}
\stoptyping
\noindentation 
The logic for determining when the content is processed is exactly the same as
for the \type|start|-\type|stop| commands.

These \type|\doif| commands each have a variant to process alternative code if
the conditions are not satisfied (like the \type|\else| branch of \type|\if|).
\starttyping
\doifmodeelse       {modes} {content} {alt}
\doifnotmodeelse    {modes} {content} {alt}
\doifallmodeselse   {modes} {content} {alt}
\doifnotallmodeselse{modes} {content} {alt}
\stoptyping

\section[sec:system] {System modes} 

Besides allowing user-definable modes, \CONTEXT\ provides some system
modes. These modes start with a \type{*} character. Here I will explain
only the more commonly used system modes; see the \CONTEXT\ modes
manual ({\tt
\hyphenatedurl{http://pragma-ade.com/general/manuals/mmodes.pdf}}) for a
complete list.

Perhaps the most useful system modes are \type|*mkii| and \type|*mkiv| which
determine whether \MKII\ or \MKIV\ is being used. These modes are handy when you
want different setups for \MKII\ and \MKIV.

Other modes are useful for very specific situations. Some of these are described
below.

A document must be run multiple times to get the cross referencing,
table of contents, etc.\
right. However, sometimes you need to do some external processing (e.g., graphic
conversion) that only needs to be done once. In such cases, the
\type|*first| mode is handy, which checks for the first run of the document. 

You can use the project-product-component structure for managing large projects
like a book series. See the \CONTEXT\
wiki article ({\tt \hyphenatedurl{http://wiki.contextgarden.net/Project_structure}})
for details of this approach. What we consider here is that both a
product and its components can be compiled on their own, 
and sometimes you want to process a component differently when it is
being compiled directly than when it is part of the complete product. This can
be checked with the modes \type|*project|, \type|*product|, \type|*component|,
and \type|*environment|. For example, the \type|*product| mode is set whenever a
product file is read; more specifically, when \type|\startproduct| is
encountered. Similarly, a mode \type|*text| is enabled when
\type|\starttext| is encountered, and likewise for the others.

A large document is typically broken down into different section blocks:
frontmatter, bodymatter, appendices, and backmatter. Internally, these section
blocks are referred to as \type|frontpart|, \type|bodypart|, \type|appendix|, and
\type|backpart|. Each section block sets a system mode with the same name. So,
if you want macros that work differently in different section blocks, you can
check for modes \type|*frontpart|, \type|*bodypart|, and so on.

\CONTEXT\ provides support for multiple languages. Languages are recognized by
their \acro{IETF} language tags, like \type|en-us| for \acro{US}
English, \type|en-gb|
for British English, \type|nl| for Dutch, \type|de| for German, etc. A document
has a main language, set with the command \type|\mainlanguage[...]|,
that is used for
translated labels like \emph{chapter} and \emph{figure}. You can also switch the
current language using \type|\language[...]| to change the hyphenation rules.
Whenever a language is chosen, its identifier is set as a mode. The mode for the main
language starts with two \type{*}. For example, when the main language
is \acro{US}
English and the current language is Dutch, the modes \type|**en-us| and
\type|*nl| are set (notice the extra \type|*| in \type|**en-us|). 

Other system modes: \type|*figure| is set when a graphic is found,
\type|*interaction| is set when interaction is enabled, \type|*grid|
is set when grid typesetting is enabled, and \type|*pdf| and \type|*dvi| 
are set depending on the whether we are generating \PDF\ or \DVI\ output. Others
are too esoteric to describe here. If you are interested, see the 
modes manual mentioned earlier.

In summary, \ConTeXt\ modes provide generalized conditional processing.
A rich set of built-in modes is available.

\StopArticle
\stoptext
